Measuring causal influence with\\ back-to-back regression

### 2019-05-23 paper/paper.tex

\begin{abstract}
Identifying causes from observations is at the core of science. This endeavor
is particularly challenging when i) potential factors are difficult to
manipulate individually and ii) observations are complex and multi-dimensional.
To address this issue, we introduce ``Back-to-Back Regression'' (B2BR), a
method designed to measure, from a set of co-varying factors, the causal influences 
that most plausible account for multidimensional observations. After proving the consistency of B2BR,
we show that our method outperforms least-squares and related
regression techniques, as well as cross-decomposition methods (e.g. canonical
correlation analysis, and partial least squares) on two tasks: causal
identification and out-of-distribution prediction. Finally, we apply B2BR to
neuroimaging recordings of 102 subjects reading word sequences. The results
show, as expected, that the early brain responses to words appear to be
specifically caused by low-level visual factors whereas late brain responses
appear to be caused by lexical factors, despite the fact that these factors
covary in the study.
% conclude  open
\end{abstract}

@@

We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
%
In summary, our model can be formalized as:
%
\begin{equation}
    Y = (XE + N)F + N',\label{eq:model}
\end{equation}
%
where $N$ and $N'$ are some unobserved noise variables.
%
While the triplet of variables $X$, $N$, and $N'$ are independent, we allow each of them to have a general covariance matrix.
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2BR) is to estimate the matrix of causal influences $E$.

@@

Back-to-Back Regression (B2BR) consists in two steps.
%
First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = \hat Y \hat G$.
%
This backward regression recovers an approximation of the causes of $Y$ in $X$.
%
Second, we estimate the linear regression coefficients $\hat H$ from $X$ to $\hat X$.
%
The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(E)$, is the desired estimate of the causal influence matrix $E$.

If using regularized least-squares, arguably the most commonly employed linear regression technique \citep{hoerl1959optimum, rifkin2007notes}, the solution of B2BR has a closed form solution:
\begin{align}
    \hat G &= (Y^\top Y + \Lambda_Y)^{-1} Y^\top X,\label{eq:solG}\\
    \hat H &=(X^\top X + \Lambda_X)^{-1} X^\top Y \hat G,\label{eq:solH}
\end{align}
%
where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices of regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{efron1992bootstrap}.
%
After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2BR prediction rule.

We summarize the B2BR procedure in Algorithm~\ref{algorithm:b2br}.
%
The rest of this section provides a theoretical guarantee on the correctness of B2BR, and an optional post-processing step to binarize the estimated causal influence matrix.

@@

On the one hand, we have leveraged B2BR to obtain the estimate $\hat{E}$, which is a diagonal matrix with real entries.
%
On the other hand, the true causal influence matrix $E$ is a binary matrix, which hard-selects the causal factors of $Y$ from $X$.
%
In this section, we provide a recipe to binarize $\hat{E}$ to estimate the collection of causal factors.

>>>>

\begin{abstract}
Identifying causes from observations is at the core of science. This endeavor
is particularly challenging when i) potential factors are difficult to
manipulate individually and ii) observations are complex and multi-dimensional.
To address this issue, we introduce ``Back-to-Back'' regression (B2B), a
method designed to measure, from a set of co-varying factors, the causal influences
that most plausibly account for multidimensional observations. After proving the consistency of B2B, we show that our method outperforms least-squares and related
regression and cross-decomposition  techniques (e.g. canonical
correlation analysis and partial least squares) on two tasks: causal
identification and out-of-distribution prediction. Finally, we apply B2B to
neuroimaging recordings of 102 subjects reading word sequences. The results
show that the early and late brain responses caused by low- and high-level
word features are more easily detected with B2B than with standard forward
regression whereas techniques.
% conclude  open
\end{abstract}

@@

We consider the measurement of multivariate signal $Y$, generated from a set of variables $X$, via some unknown linear apparatus $F$.
%
Not all the variables in $X$ exert a causal influence on $Y$.
%
By considering a square binary diagonal matrix of \emph{causal influences} $E$, we denote by $XE$ the causal factors of $Y$.
%
In summary, our model can be formalized as:
%
\begin{equation}
    Y = (XE + N)F + N',\label{eq:model}
\end{equation}
%
where $N$ and $N'$ are some unobserved noise variables.
%
While the triplet of variables $X$, $N$, and $N'$ are independent, we allow each of them to have a general covariance matrix.
%
In practice, we observe $n$ samples $(X, Y)$ from the model.
%
This process, along with the sizes of all variables involved, is illustrated in Figure~\ref{fig:b2b}.
%
Given the model in Equation~\eqref{eq:model}, \textbf{the goal} of Back-to-Back Regression (B2B) is to estimate the matrix of causal influences $E$.

@@

Back-to-Back Regression (B2B) consists in two steps.
%
First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = \hat Y \hat G$.
%
This backward regression recovers an approximation of the causes of $Y$ in $X$.
%
Second, we estimate the linear regression coefficients $\hat H$ from $X$ to $\hat X$.
%
The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(E)$, is the desired estimate of the causal influence matrix $E$.

If using regularized least-squares, arguably the most commonly employed linear regression technique \citep{hoerl1959optimum, rifkin2007notes}, the solution of B2B has a closed form solution:
\begin{align}
    \hat G &= (Y^\top Y + \Lambda_Y)^{-1} Y^\top X,\label{eq:solG}\\
    \hat H &=(X^\top X + \Lambda_X)^{-1} X^\top Y \hat G,\label{eq:solH}
\end{align}
%
where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices of regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2B is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split \citep{efron1992bootstrap}.
%
After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2B prediction rule.

We summarize the B2B procedure in Algorithm~\ref{algorithm:b2br}.
%
The rest of this section provides a theoretical guarantee on the correctness of B2B, and an optional post-processing step to binarize the estimated causal influence matrix.

@@

On the one hand, we have leveraged B2B to obtain the estimate $\hat{E}$, which is a diagonal matrix with real entries.
%
On the other hand, the true causal influence matrix $E$ is a binary matrix, which hard-selects the causal factors of $Y$ from $X$.
%
In this section, we provide a recipe to binarize $\hat{E}$ to estimate the collection of causal factors.
